A Way Forward in Parallelising Dynamic Languages

### 2014-06-17 talk/icooolps2014/position-paper.tex
second round of feedback

\begin{table}[!ht]
  \begin{center}
    \begin{tabular}{|l|l|}
      \hline
      Thread 1 & Thread 2 \\
      \hline
      \multicolumn{2}{|l|}{\texttt{A = B = 0}} \\
      \hline
      \texttt{A = 1} & \texttt{B = 1}\\
      \texttt{if B == 0:} & \texttt{if A == 0:}\\
      \texttt{    critical section} & \texttt{    critical section}\\
      \hline
    \end{tabular}
    \caption{Critical section with a sequential consistency model.}
    \label{tab:seq_cons}
  \end{center}
\end{table}

@@

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronise memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is
known that using locks for synchronisation can be hard at
times~\cite{christopher10,victor11,shan08}. They are non-composable,
have overhead, may deadlock, limit scalability, and add to the overall
complexity of the program logic. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronisation mechanism called \emph{atomic
  blocks}~\cite{tim03,tim05}.

@@

\begin{table*}[ht]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & \textbf{GIL} & \textbf{Fine-grained locking}
    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
    \hline
    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
    \hline
    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
    \hline
    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
    \hline
    Better synchronisation        & o    & o  &      & o  & ++   \\
    \hline
    Implementation                & ++   & -  & ++   & ++ & ++   \\
    \hline
    External libraries            & ++   & ++ & ++   & ++ & ++   \\
    \hline
  \end{tabular}
  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
  \label{tab:comparison}
\end{table*}

\subsection{Potential Solutions}

@@

In the end, fine-grained locking can transparently replace the GIL and
therefore parallelise existing applications, generally without any
changes\footnote{There are rare cases where not having atomic
  bytecodes actually changes the semantics. E.g.\ in Jython,
  \texttt{dict1.update(dict2)} is not atomic: it first reads data from
  \texttt{dict2} with \texttt{dict2}'s lock, and then puts it into
  \texttt{dict1} with \texttt{dict1}'s lock. A lot can happen
  in-between.}. An implementation has to follow the GIL semantics very
closely, otherwise it may expose some latent data races in existing
applications which are just not exposed with a GIL. This approach does
however not provide a better parallelising synchronisation mechanism
to the application like e.g. atomic blocks.

%% - support of atomic blocks?\\
%% - hard to get right (deadlocks, performance, lock-granularity)\\
%% - very hard to get right for a large language\\
%% - hard to retro-fit, as all existing code assumes GIL semantics\\
%% - (there are some semantic differences, right? not given perfect lock-placement, but well)
%% ( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )

@@

There are also approaches that work around the GIL instead of trying
to replace it. If an application can be split into completely
independent parts that only very rarely need to share anything, or
only do so via an external program like a database, then it is
sensible to have one GIL per independent part. At the extreme, there
are applications that parallelise perfectly simply by running
independent processes; some web servers and some numeric computations
do. We will consider here a slightly more general approach: the
\emph{multiprocessing}\footnote{https://docs.python.org/2/library/multiprocessing.html}
module of Python. In essence, it uses process-forking to provide the
application with multiple interpreters that can run in parallel.
Communication is then done explicitly through pipes.

@@

%% - often needs major restructuring of programs (explicit data exchange)\\
%% - sometimes communication overhead is too large\\
%% - shared memory is a problem, copies of memory are too expensive

@@

Following the above argumentation for each approach, we assembled a
general overview in Table \ref{tab:comparison}. The general picture is
everything else than clear. It looks like HTM may be a good solution
to replace the GIL in the near future. Current implementations are
however far too limiting and do not provide good scaling.

Allowing for parallel execution just means that dynamic languages
catch up to all other languages that already provide real
parallelism. This is why we think that only the STM approach is a
viable solution in the long-term. It provides the application with a
simple memory model (sequential consistency) and a composable way to
synchronise memory accesses using atomic blocks.

@@

We believe that further work to reduce the overhead of STM is
very worthwhile. In fact, considering some analogies that have been
drawn between garbage collection and transactional memory~\cite{dan07},
we believe that it is worthwhile to focus the STM research more
specifically onto the context shown in this paper --- for use in
implementations of high-level languages, rather than as a tool
directly used by the programmer.

@@

>>>>

\begin{table}[!ht]
  \begin{center}
    \begin{tabular}{|l|l|}
      \hline
      Thread 1                      & Thread 2                      \\
      \hline
      \multicolumn{2}{|l|}{\texttt{A = B = 0}}                      \\
      \hline
      \texttt{A = 1}                & \texttt{B = 1}                \\
      \texttt{if B == 0:}           & \texttt{if A == 0:}           \\
      \texttt{    critical section} & \texttt{    critical section} \\
      \hline
    \end{tabular}
    \caption{Critical section with a sequential consistency model.}
    \label{tab:seq_cons}
  \end{center}
\end{table}

@@

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronise memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is
known that using locks for synchronisation can be hard at
times~\cite{christopher10,victor11,shan08}. They are non-composable,
have overhead, may deadlock, limit scalability, and add to the overall
complexity of the program logic. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronisation mechanism called \emph{atomic
  blocks}~\cite{tim03,tim05}. This is also suggested by
\cite{christopher10,victor11} as an easier mechanism than locks.

@@

\begin{table*}[ht]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & \textbf{GIL} & \textbf{Fine-grained locking}
    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
    \hline
    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
    \hline
    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
    \hline
    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
    \hline
    Better synchronisation        & o    & o  & +    & o  & ++   \\
    \hline
    Implementation                & ++   & -  & ++   & ++ & ++   \\
    \hline
    External libraries            & ++   & ++ & ++   & ++ & ++   \\
    \hline
  \end{tabular}
  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
  \label{tab:comparison}
\end{table*}

\subsection{Potential Solutions}
\label{sec:pot_solutions}

@@

In the end, fine-grained locking can transparently replace the GIL and
therefore parallelise existing applications, generally without any
changes\footnote{There are rare cases where not having atomic
  bytecodes actually changes the semantics. E.g.\ in Jython,
  \texttt{dict1.update(dict2)} is not atomic: it first reads data from
  \texttt{dict2} with \texttt{dict2}'s lock, and then puts it into
  \texttt{dict1} with \texttt{dict1}'s lock. A lot can happen
  in-between.}. An implementation has to follow the GIL semantics very
closely, otherwise it may expose some latent data races in existing
applications which are just not exposed with a GIL. This approach does
however not provide a better parallelising synchronisation mechanism
to the application like e.g.\ atomic blocks.

@@

There are also approaches that work around the GIL instead of trying
to replace it. If an application can be split into completely
independent parts that only very rarely need to share anything, or
only do so via an external program like a database, then it is
sensible to have one GIL per independent part. At the extreme, there
are applications that parallelise perfectly simply by running
independent processes; some web servers and some numeric computations
do.

We will consider here a slightly more general approach: the
\emph{multiprocessing}~\cite{multiprocessing}
module of Python. In essence, it uses process-forking to provide the
application with multiple interpreters that can run in parallel.
Communication is then done explicitly through pipes.\footnote{There
are multiple alternative designs like e.g.\ actors or tuple spaces.
Since they are similar and do not replace the GIL directly, we
focus on the example of \emph{multiprocessing}.}

@@

@@

Following the above argumentation for each approach, we assembled a
general overview in Table \ref{tab:comparison}. The points were assigned
according to the criteria described in \ref{sec:pot_solutions}. Since
the criteria are defined intuitively, there are no formal justifications
for the assigned points. The reader is thus advised to take them with a
grain of salt.

The general picture is everything else than clear. It looks like HTM
may be a good solution to replace the GIL in the near future. Current
implementations are however far too limiting, not widely available,
and do not provide good scaling.

Allowing for parallel execution just means that dynamic languages
catch up to all other languages that already provide real
parallelism. This is why we think that only the STM approach is a
viable solution in the long-term. It unifies both, the simple memory
model (sequential consistency) and the synchronisation of memory accesses
using composable atomic blocks. It is not \emph{only} a simple GIL
replacement.

@@

We think that further work to reduce the overhead of STM is
very worthwhile. In fact, considering some analogies that have been
drawn between garbage collection and transactional memory~\cite{dan07},
we think that it is worthwhile to focus the STM research more
specifically onto the context shown in this paper --- for use in
implementations of high-level languages, rather than as a tool
directly used by the programmer.

@@

\bibitem{multiprocessing}
  The Multiprocessing Module of Python,
  \url{docs.python.org/2/library/multiprocessing.html}
